Novel risk loci for COVID-19 hospitalization among admixed American populations

Version 1

Abstract

Abstract

The genetic basis of severe COVID-19 has been thoroughly studied and many genetic risk factors shared between populations have been identified. However, reduced sample sizes from non-European groups have limited the discovery of population-specific common risk loci. In this second study nested in the SCOURGE consortium, we have conducted the largest GWAS meta-analysis for COVID-19 hospitalization in admixed Americans, comprising a total of 4,702 hospitalized cases recruited by SCOURGE and other seven participating studies in the COVID-19 Host Genetic Initiative. We identified four genome-wide significant associations, two of which constitute novel loci and first discovered in Latin-American populations (BAZ2B and DDIAS). A trans-ethnic meta-analysis revealed another novel cross-population risk locus in CREBBP. Finally, we assessed the performance of a cross-ancestry polygenic risk score in the SCOURGE admixed American cohort.

Introduction

To date, more than 50 loci associated to COVID-19 susceptibility, hospitalization, and severity have been identified using genome-wide association studies (GWAS)1,2. The COVID-19 Host Genetics Initiative (HGI) has made significant efforts3 to augment the power to identify disease loci by recruiting individuals from diverse populations and conducting a trans-ancestry meta-analysis. Despite this, the lack of genetic diversity and a focus on cases of European ancestries still predominate in the studies4,5. Besides, while trans-ancestry meta-analyses are a powerful approach for discovering shared genetic risk variants with similar effects across populations6, they may fail to identify risk variants that have larger effects on particular underrepresented populations. Genetic disease risk has been shaped by the particular evolutionary history of populations and the environmental exposures7. Their action is particularly important for infectious diseases due to the selective constrains that are imposed by the host-pathogen interactions8,9. Literature examples of this in COVID-19 severity includes a DOCK2 gene variant in East Asians10, and frequent loss of function variants in IFNAR1 and IFNAR2 genes in Polynesian and Inuit populations, respectively11,12.

Including diverse populations in case-control GWAS studies with unrelated participants usually require a prior classification of individuals in genetically homogeneous groups, which are typically analysed separately to control the population stratification effects13. Populations with recent admixture impose an additional challenge to the GWAS due to their complex genetic diversity and linkage disequilibrium (LD) patterns, requiring the development of alternative approaches and a careful inspection of results to reduce the false positives due to population structure7. In fact, there are benefits in study power from modelling the admixed ancestries either locally, at regional scale in the chromosomes, or globally, across the genome, depending on factors such as the heterogeneity of the risk variant in frequencies or the effects among the ancestry strata14. Despite the development of novel methods specifically tailored for the analysis of admixed populations15, the lack of a standardized analysis framework and the difficulties to confidently cluster the admixed individuals into particular genetic groups often leads to their exclusion from GWAS.

The Spanish Coalition to Unlock Research on Host Genetics on COVID-19 (SCOURGE) recruited COVID-19 patients between March and December 2020 from hospitals across Spain and from March 2020 to July 2021 in Latin-America (https://www.scourge-covid.org). A first GWAS of COVID-19 severity among Spanish patients of European descent revealed novel disease loci and explored age and sex varying effects of the genetic factors16. Here we present the findings of a GWAS meta-analysis in admixed American (AMR) populations, comprising individuals from the SCOURGE Latin-American cohort and the HGI studies, which allowed to identify two novel severe COVID-19 loci, BAZ2B and DDIAS. Further analyses modelling the admixture from three genetic ancestral components and performing a trans-ethnic meta-analysis led to the identification of an additional risk locus near CREBBP. We finally assessed a cross-ancestry polygenic risk score model with variants associated with critical COVID-19.

Results

Meta-analysis of COVID-19 hospitalization in admixed Americans

Study cohorts

Within the SCOURGE consortium, we included 1,608 hospitalized cases and 1,887 controls (not hospitalized COVID-19 patients) from Latin-American countries and from recruitments of individuals of Latin-American descend conducted in Spain (Supplementary Table 1). Quality control details and estimation of global genetic inferred ancestry (GIA) (supplementary Figure 1) are described in Methods, whereas clinical and demographic characteristics of patients included in the analysis are shown in Table 1. Summary statistics from the SCOURGE cohort were obtained under a logistic mixed model with the SAIGE model (Methods). Another seven studies participating in the COVID-19 HGI consortium were included in the meta-analysis of COVID-19 hospitalization in admixed Americans (Figure 1).

Flow chart of this study.

GWAS meta-analysis

We performed a fixed-effects GWAS meta-analysis using the inverse of the variance as weights for the overlapping markers. The combined GWAS sample size consisted of 4,702 admixed AMR hospitalized cases and 68,573 controls.

This GWAS meta-analysis revealed genome-wide significant associations at four risk loci (Table 2, Figure 2), two of which (BAZ2B and DDIAS) were novel discoveries. Variants of these loci were prioritized by positional and expression quantitative trait loci (eQTL) mapping with FUMA, identifying four lead variants linked to other 310 variants and 31 genes (Supplementary Tables 2-4). A gene-based association test revealed a significant association in BAZ2B and in previously known COVID-19 risk loci: LZTFL1, XCR1, FYCO1, CCR9, and IFNAR2 (Supplementary Table 5).

A) Manhattan plot for the admixed AMR GWAS meta-analysis. Probability thresholds at p=5×10-8 and p=5×10-5 are indicated by the horizontal lines. Genome-wide significant associations with COVID-19 hospitalizations were found in chromosome 2 (within BAZ2B), chromosome 3 (within LZTFL1), chromosome 6 (within FOXP4), and chromosome 11 (within DDIAS). A Quantile-Quantile plot is shown in supplementary Figure 2. B) Regional association plots for rs1003835 at chromosome 2 and rs77599934 at chromosome 11; C) Allele frequency distribution across The 1000 Genomes Project populations for the lead variants rs1003835 and rs77599934.

Located within the gene BAZ2B, the sentinel variant rs13003835 is an intronic variant associated with an increased risk of COVID-19 hospitalization (Odds Ratio [OR]=1.20, 95% Confidence Interval [CI]=1.12-1.27, p=3.66×10-8). This association was not previously reported in any GWAS of COVID-19 published to date. Interestingly, rs13003835 did not reach significance (p=0.972) in the COVID-19 HGI trans-ancestry meta-analysis including the five population groups1. Based on our mapping strategy (see Methods), we also prioritized PLA2R1, LY75, WDSUB1, and CD302 in this locus.

The other novel risk locus is led by the sentinel variant rs77599934, a rare intronic variant located in chromosome 11 within DDIAS and associated with risk of COVID-19 hospitalization (OR=2.27, 95%CI=1.70-3.04, p=2.26×10-8). The PRCP gene was an additional prioritized gene at this locus.

We also observed a suggestive association with rs2601183 in chromosome 15, which is located between ZNF774 and IQGAP1 (allele-G OR=1.20, 95%CI=1.12-1.29, p=6.11×10-8, see Supplementary Table 2), which has not yet been reported in any other GWAS of COVID-19 to date. This sentinel variant is in perfect LD (r2=1) with rs601183, an eQTL of ZNF774 in the lung.

The GWAS meta-analysis also pinpointed two significant variants at known loci, LZTFL1 and FOXP4. The SNP rs35731912 was previously associated with COVID-19 severity in EUR populations17, and it was mapped to LZTFL1. As for rs2477820, while it is a novel risk variant within gene FOXP4, it has a moderate LD (r2=0.295) with rs2496644, which has been linked to COVID-19 hospitalization18. This is consistent with the effects of LD in tag-SNPs when conducting GWAS in diverse populations.

Functional mapping of novel risk variants

Bayesian fine mapping

We performed different approaches to narrow down the prioritized loci to a set of most probable genes driving the associations. First, we computed credible sets at the 95% confidence for causal variants and annotated them with VEP and the V2G aggregate scoring (Supplementary Table 6, Supplementary Figure 3). The 95% confidence credible set from the region of chromosome 2 around rs13003835 included 76 variants. However, the approach was unable to converge allocating variants in a 95% confidence credible set for the region in chromosome 11.

Colocalization of eQTLs

To determine if the novel genetic risk loci were associated with gene expression in relevant tissues (whole blood, lung, lymphocytes, and oesophagus mucosa), we computed the posterior probabilities (PP) of colocalization for overlapping variants allocated to the 95% confidence credible set. We used the GTEx v8 tissues as the main expression dataset, although it is important to consider that the eQTL associations were carried out mainly on individuals of EUR ancestries. To confirm the colocalization in other ancestries, we also performed analyses on three expression datasets computed on admixed AMR, leveraging data from individuals with high African GIA, high Native-American ancestry, and from a pooled cohort (Methods). Results are shown in the supplementary Table 7.

Five genes (LY75, BAZ2B, CD302, WDSUB1, and PLA2R1) were the candidates for eQTL colocalization in the associated region in chromosome 2. However, LY75 emerged as the most likely causal gene for this locus since the colocalization in whole blood was supported with a PP for H4 (PPH4) of 0.941 and with robust results (supplementary Figure 4). Moreover, this also allowed to prioritize rs12692550 as the most probable causal variant for both traits at this locus with a PP_SNP_H4 of 0.74. Colocalization with gene expression data from admixed AMR validated this finding. LY75 also had evidence of colocalization in lungs (PPH4=0.887) and the esophagus mucosa (PPH4=0.758). However, we could not prioritize a single causal variant in these two other tissues and sensitivity analyses revealed a weak support.

CD302 and BAZ2B were the second and third most likely genes that could drive the association, respectively, according to the colocalization evidence. CD302 was the most probable according to the high AFR genetic ancestries dataset (supplementary Figure 5).

Despite the chromosome 11 region failing to colocalize with gene expression associations for any of the tissues, the lead variant rs77599934 is in moderate-to-strong LD (r2=0.776) with rs60606421, which is an eQTL associated to a reduced expression of DDIAS in the lungs (supplementary Figure 6). The highest PPH4 for DDIAS was in the high AFR genetic ancestry expression dataset (0.71).

Transcriptome-wide association study (TWAS)

Five novel genes, namely SLC25A37, SMARCC1, CAMP, TYW3, and S100A12 (supplementary Table 8) were found significantly associated in the cross-tissue TWAS. To our knowledge, these genes have not been reported previously in any COVID-19 TWAS or GWAS analyses published to date. In the single tissue analyses, ATP5O and CXCR6 were significantly associated in lungs, CCR9 was significantly associated in whole blood, and IFNAR2 and SLC25A37 were associated in lymphocytes.

Likewise, we carried out the TWAS analyses using the models trained in the admixed populations. However, no significant gene-pairs were detected in this case. The 50 genes with the lowest p-values are shown in the supplementary Table 9.

Genetic architecture of COVID-19 hospitalization in AMR populations

Allele frequencies of rs13003835 and rs77599934 across ancestries

Neither rs13003835 (BAZ2B) or rs77599934 (DDIAS) were significantly associated in the COVID-19 HGI B2 cross-population or population-specific meta-analyses. Thus, we investigated their allele frequencies (AF) across populations and compared their effect sizes.

According to gnomAD v3.1.2, the T allele at rs13003835 (BAZ2B) has an AF of 43% in admixed AMR groups while AF is lower in the EUR populations (16%) and in the global sample (29%). Local ancestry inference (LAI) reported by gnomAD shows that within the Native-American component, the risk allele T is the major allele, whereas it is the minor allele within the African and European LAI components. These large differences in AF might be the reason underlying the association found in AMR populations. However, when comparing effect sizes between populations, we found that they were in opposite direction between SAS-AMR and EUR-AFR-EAS and that there was a large heterogeneity among them (Figure 3).

Forest plot showing effect sizes and the corresponding confidence intervals for the sentinel variants identified in the AMR meta-analysis across populations. All beta values with their corresponding CIs were retrieved from the B2 population-specific meta-analysis from the HGI v7 release, except for AMR, for which the beta value and IC from the HGIAMR-SCOURGE meta-analysis is represented.

rs77599934 (DDIAS) had an AF of 1.1% for the G allele in the non-hospitalized controls (Table 2), in line with the recorded gnomAD AF of 1% in admixed AMR groups. This variant has potential to be population-specific variant, given the allele frequencies in other population groups such as EUR (0% in Finnish, 0.025% in non-Finnish), EAS (0%) and SAS (0.042%) and its greater effect size over AFR populations (Figure 3). Examining the LAI, the G allele occurs at 1.1% frequency in the African component while it is almost absent in the Native-American and European. Due to its low MAF, rs77599934 was not analyzed in the COVID-19 HGI B2 cross-population meta-analysis and was only present in the HGI B2 AFR population-specific meta-analysis, precluding the comparison (Figure 3). For this reason, we retrieved the variant with the lowest p-value within a 50 kb region around rs77599934 in the COVID-19 HGI cross-population analysis to investigate if it was in moderate-to-strong LD with our sentinel variant. The variant with the smallest p-value was rs75684040 (OR=1.07, 95%CI=1.03-1.12, p=1.84×10-3). Yet, LD calculations using the 1KGP phase 3 dataset indicated that rs77599934 and rs75684040 were poorly correlated (r2=0.11).

Cross-population meta-analyses

We carried out two cross-ancestry inverse variance-weighted fixed-effects meta-analyses with the admixed AMR GWAS meta-analysis results to evaluate whether the discovered risk loci replicated when considering other population groups. In doing so, we also identified novel cross-population COVID-19 hospitalization risk loci.

First, we combined the SCOURGE Latin American GWAS results with the HGI B2 ALL analysis (supplementary Table 10). We refer to this analysis as the SC-HGIALL meta-analysis. Out of the 40 genome-wide significant loci associated with COVID-19 hospitalization in the last HGI release1, this study replicated 39 and the association was stronger than in the original study in 29 of those (supplementary Table 11). However, the variant rs13003835 located in BAZ2B did not replicate (OR=1.00, 95%CI=0.98-1.03, p=0.644).

In this cross-ancestry meta-analysis, we replicated two associations that were not found in HGIv7 albeit they were sentinel variants in the latest GenOMICC meta-analysis2. We found an association at the CASC20 locus led by the variant rs2876034 (OR=0.95, 95%CI=0.93-0.97, p=2.83×10-8). This variant is in strong LD with the sentinel variant of that study (rs2326788, r2=0.92), which was associated with critical COVID-192. Besides, this meta-analysis identified the variant rs66833742 near ZBTB7A associated with COVID-19 hospitalization (OR=0.94, 95%CI=0.92-0.96, p=2.50×10-8). Notably, rs66833742 or its perfect proxy rs67602344 (r2=1) are also associated with upregulation of ZBTB7A in whole blood and in esophagus mucosa. This variant was previously associated with COVID-19 hospitalization2.

In a second analysis, we also explored the associations across the defined admixed AMR, EUR, and AFR ancestral sources by combining through meta-analysis the SCOURGE Latin American GWAS results with the HGI studies in EUR, AFR, and admixed AMR, and excluding those from EAS and SAS (Supplementary Table 12). We refer to this as the SC-HGI3POP meta-analysis. The association at rs13003835 (BAZ2B, OR=1.01, 95%CI=0.98-1.03, p=0.605) was not replicated and rs77599934 near DDIAS could not be assessed, although the association at the ZBTB7A locus was confirmed (rs66833742, OR=0.94, 95%CI=0.92-0.96, p=1.89×10-8). The variant rs76564172 located near CREBBP also reached statistical significance (OR=1.31, 95% CI=1.25-1.38, p=9.64×10-9). The sentinel variant of the region linked to CREBBP (in the trans-ancestry meta-analysis) was also subjected a Bayesian fine mapping (supplementary Table 6) and colocalization with eQTLs under the GTEx v8 MASHR models in lungs, esophagus mucosa, whole blood, and transformed lymphocytes. Eight variants were included in the credible set for the region in chromosome 16 (meta-analysis SC-HGI3POP), although CREBBP did not colocalize in any of the tissues.

Polygenic risk score models

Using the 49 variants associated with disease severity that are shared across populations according to the HGIv7, we constructed a polygenic risk score (PGS) model to assess its generalizability in the admixed AMR (Supplementary Table 13). First, we calculated the PGS for the SCOURGE Latin Americans and explored the association with COVID-19 hospitalization under a logistic regression model. The PGS model was associated with a 1.48-fold increase in COVID-19 hospitalization risk per every PGS standard deviation. It also contributed to explain a slightly larger variance (R2=1.07%) than the baseline model.

Subsequently, we divided the individuals into PGS deciles and percentiles to assess their risk stratification. The median percentile among controls was 40, while in cases it was 63. Those in the top PGS decile exhibited a 5.90-fold (95% CI=3.29-10.60, p=2.79×10-9) greater risk compared to individuals in the lowest decile, whereas the effects for the rest of the comparisons were much milder.

We also examined the distribution of PGS scores across a 5-level severity scale to further determine if there was any correspondence between clinical severity and genetic risk. Median PGS scores were lower in the asymptomatic and mild groups, whereas higher median scores were observed in the moderate, severe, and critical patients (Figure 4). We fitted a multinomial model using the asymptomatic class as reference and calculated the OR for each category (Supplementary Table 13), observing that the disease genetic risk was similar among asymptomatic, mild, and moderate patients. Given that the PGS was built with variants associated with critical disease and/or hospitalization and that the categories severe and critical correspond to hospitalized patients, these results underscore the ability of cross-ancestry PGS for risk stratification even in an admixed population.

(A) Polygenic risk stratified by PGS deciles comparing each risk group against the lowest risk group (OR-95%CI); (B) Distribution of the PGS scores in each of the severity scale classes (0-Asymptomatic, 1-Mild disease, 2-Moderate disease, 3-Severe disease, 4-Critical disease).

Finally, we incorporated the novel lead SNPs from our AMR meta-analysis (rs13003835, rs2477820, and rs77599934) into the PGS model. Their inclusion in the model contributed to explain a larger variance (R2=1.74%) than the model without them. This result, however, should be taken with caution given the risk of overfitting due to the use of the same subjects both for the derivation and testing of the variants.

Discussion

We have conducted the largest GWAS meta-analysis of COVID-19 hospitalization in admixed AMR to date. While the genetic risk basis discovered for COVID-19 is largely shared among populations, trans-ancestry meta-analyses on this disease have primarily included EUR samples. This dominance of GWAS in Europeans, and the subsequent bias in sample sizes, can mask population-specific genetic risks (i.e., variants that are monomorphic in some populations) or be less powered to detect risk variants having higher allele frequencies in population groups other than Europeans. In this sense, after combining data from admixed AMR patients, we found two risk loci which are first discovered in a GWAS of Latin-American populations. Interestingly, the sentinel variant rs77599934 in the DDIAS gene is a rare coding variant (∼1% for allele G) with a large effect on COVID-19 hospitalization that is nearly monomorphic in most of the other populations. This has likely led to its exclusion from the cross-populations meta-analyses conducted to date, remaining undetectable.

Fine mapping of the region harbouring DDIAS did not reveal further information about which gene could be the more prone to be causal, or about the functional consequences of the risk variant. However, DDIAS, known as damage-induced apoptosis suppressor gene, is itself a plausible candidate gene. It has been linked to DNA damage repair mechanisms: research showed that depletion of DDIAS led to an increase of ATM phosphorylation and the formation of p53-binding protein (53BP1) foci, a known biomarker of DNA double-strand breaks, suggesting a potential role in double-strand break repair19. Similarly, elevated levels of phosphorylated nuclear histone 2AXγ were detected after knocking down DDIAS, further emphasizing its role in DNA damage20. Interestingly, a study found that the infection by SARS-CoV-2 also triggered the phosphorylation of the ATM kinase and inhibited repair mechanisms, causing the accumulation of DNA damage21. This same study reported the activation of the pro-inflammatory pathway p38/MAPK by the virus, which was as well prompted after knocking-down DDIAS20.

Regarding lung function, the role of DDIAS in lung cancer has been widely studied. It has been proposed as a potential biomarker for lung cancer after finding that it interacts with STAT3 in lung cancer cells, regulating IL-622,23 and thus mediating inflammatory processes. Furthermore, another study determined that its blockade inhibited lung cancer cell growth20. The sentinel variant was in strong LD with an eQTL that reduced gene expression of DDIAS in lung, and our findings suggest that DDIAS gene may be indeed involved in viral response. Hence, one reasonable hypothesis is that reduced expression of DDIAS could potentially facilitate SARS-CoV-2 infection through the downregulation of pathways involved in DNA repairment and inflammation. Another prioritized gene from this region was PRCP, an angiotensinase that has been linked to hypertension and for which a hypothesis on its role on COVID-19 progression has been raised24,25.

The risk region found in chromosome 2 prioritized more than one gene. The lead variant rs13003835 is located within BAZ2B. BAZ2B encodes one of the regulatory subunits of the Imitation switch (ISWI) chromatin remodelers26 constituting the BRF-1/BRF-5 complexes with SMARCA1 and SMARCA5, respectively, and the association signal colocalized with eQTLs in whole blood. The gene LY75 (encoding the lymphocyte antigen 75) also colocalized with eQTLs in whole blood, esophagus mucosa, and lung tissues. Lymphocyte antigen 75 is involved in immune processes through antigen presentation in dendritic cells and endocytosis27, and has been associated with inflammatory diseases, representing also a compelling candidate for the region. Increased expression of LY75 has been detected within hours after the infection by SARS-CoV-228,29. Lastly, the signal of CD302 colocalized in individuals with high AFR ancestral admixture in whole blood. This gene is located in the vicinity of LY75 and both conform the readthrough LY75-CD302. It is worth noting that differences in AF for this variant suggest that analyses in AMR populations might be more powered to detect the association, supporting the necessity of population-specific studies.

A third novel risk region was observed in chromosome 15, between the genes IQGAP1 and ZNF774, although not reaching genome-wide significance.

Secondary analyses revealed five TWAS-associated genes, some of which have been already linked to severe COVID-19. In a comprehensive multi-tissue gene expression profiling study30, decreased expression of CAMP and S100A8/S100A9 genes in COVID-19 severe patients was observed, while another study detected the upregulation of SCL25A37 among severe COVID-19 patients31. SMARCC1 is a subunit of the SWI/SNF chromatin remodelling complex that has been identified as pro-viral for SARS-CoV-2 and other coronavirus strains through a genome-wide screen32. This complex is crucial for ACE2 expression and the viral entry in the cell33.

To explore the genetic architecture of the trait among admixed AMR populations, we performed two cross-ancestry meta-analyses including the SCOURGE Latin-American cohort GWAS findings. We found that the two novel risk variants did not associate with COVID-19 hospitalization outside the population-specific meta-analysis, highlighting the importance of complementing trans-ancestry meta-analyses with group-specific analyses. Notably, this analysis did not replicate the association at the DSTYK locus, which was associated with severe COVID-19 in Brazilian individuals with higher European admixture34. This lack of replication supports the initial hypothesis of that study suggesting that the risk haplotype derived from European populations, as we have reduced the weight of this ancestral contribution in our study by excluding those individuals.

Moreover, these cross-ancestry meta-analyses pointed to three loci that were not genome-wide significant in the HGIv7 ALL meta-analysis: a novel locus at CREBBP, and two loci at ZBTB7A and CASC20 that were reported in another meta-analysis. CREBBP and ZBTB7A achieved a stronger significance when considering only EUR, AFR, and admixed AMR GIA groups. According to a recent study, elevated levels of the ZBTB7A gene promote a quasi-homeostatic state between coronaviruses and host cells, preventing cell death by regulating oxidative stress pathways35. This gene is involved in several signalling pathways, such as B and T cell differentiation36. On a separate note, CREBBP encodes the CREB binding protein (CBP), involved in transcription activation, that is known to positively regulate the type I interferon response through virus-induced phosphorylation of IRF-337. Besides, the CREBP/CBP interaction has been implicated in SARS-CoV-2 infection38 via the cAMP/PKA pathway. In fact, cells with suppressed CREBBP gene expression exhibit reduced replication of the so called Delta and Omicron SARS-CoV-2 variants38.

The cross-population PGS model effectively stratified individuals based on their genetic risk and demonstrated consistency with the clinical severity classification of the patients. The inclusion of the new variants in the PGS model slightly improved the predictive value of the PGS. However, it is important to confirm this last finding in an external admixed AMR cohort to address potential overfitting arising from using the same individuals both for the discovery of the associations and for testing the model.

This study is subject to limitations, mostly concerning the sample recruitment and composition. The SCOURGE Latino-American sample size is small and the GWAS is underpowered. Another limitation is the difference in case-control recruitment across sampling regions that, yet controlled for, may reduce the ability to observe significant associations driven by different compositions of the populations. In this sense, the identified risk loci might not replicate in a cohort lacking any of the parental population sources from the three-way admixture. Likewise, we could not explicitly control for socio-environmental factors that could have affected COVID-19 spread and hospitalization rates, although genetic principal components are known to capture non-genetic factors. Finally, we must acknowledge the lack of a replication cohort. We have used all the available GWAS data for COVID-19 hospitalization in admixed AMR in this meta-analysis due to the low number of studies conducted. Therefore, we had no studies to replicate or validate the results. These concerns may be addressed in the future by including more AMR GWAS studies in the meta-analysis, both by involving diverse populations in study designs and by supporting research from countries in Latin-America.

This study provides novel insights into the genetic basis of COVID-19 severity, emphasizing the importance of considering host genetic factors through using non-European populations, especially of admixed sources. Such complementary efforts can pin down new variants and increase our knowledge on the host genetic factors of severe COVID-19.

Materials and methods

GWAS in Latin Americans from SCOURGE

The SCOURGE Latin American cohort

A total of 3,729 of COVID-19 positive cases were recruited across five countries from Latin America (Mexico, Brazil, Colombia, Paraguay, and Ecuador) by 13 participating centres (supplementary Table 1) from March 2020 to July 2021. In addition, we included 1,082 COVID-19 positive individuals recruited between March and December 2020 in Spain who either had evidence of origin from a Latin American country or showed inferred genetic admixture between AMR, EUR, and AFR (with < 0.05% SAS/EAS). These individuals were excluded from a previous SCOURGE study that focused on participants with European genetic ancestries16. We used hospitalization as a proxy for disease severity and defined as cases those COVID-19 positive patients that underwent hospitalization as a consequence of the infection and used as controls those that did not need hospitalization due to COVID-19.

Samples and data were collected with informed consent after the approval of the Ethics and Scientific Committees from the participating centres and by the Galician Ethics Committee Ref 2020/197. Recruitment of patients from IMSS (in Mexico, City), was approved by of the National Comitte of Clinical Research, from Instituto Mexicano del Seguro Social, Mexico (protocol R-2020-785-082).

Samples and data were processed following normalized procedures. The REDCap electronic data capture tool39,40, hosted at Centro de Investigación Biomédica en Red (CIBER) from the Instituto de Salud Carlos III (ISCIII), was used to collect and manage demographic, epidemiological, and clinical variables. Subjects were diagnosed for COVID-19 based on quantitative PCR tests (79.3%), or according to clinical (2.2%) or laboratory procedures (antibody tests: 16.3%; other microbiological tests: 2.2%).

SNP array genotyping

Genomic DNA was obtained from peripheral blood and isolated using the Chemagic DNA Blood 100 kit (PerkinElmer Chemagen Technologies GmbH), following the manufacturer’s recommendations.

Samples were genotyped with the Axiom Spain Biobank Array (Thermo Fisher Scientific) following the manufacturer’s instructions in the Santiago de Compostela Node of the National Genotyping Center (CeGen-ISCIII; http://www.usc.es/cegen). This array contains probes for genotyping a total of 757,836 SNPs. Clustering and genotype calling were performed using the Axiom Analysis Suite v4.0.3.3 software.

Quality control steps and variant imputation

A quality control (QC) procedure using PLINK 1.941 was applied to both samples and the genotyped SNPs. We excluded variants with a minor allele frequency (MAF) <1%, a call rate <98%, and markers strongly deviating from Hardy-Weinberg equilibrium expectations (p<1×10-6) with mid-p adjustment. We also explored the excess of heterozygosity to discard potential cross-sample contaminations. Samples missing >2% of the variants were filtered out. Subsequently, we kept the autosomal SNPs and removed high LD regions and conducted LD-pruning (windows of 1,000 SNPs, with step size of 80 and r2 threshold of 0.1) to assess kinship and estimate the global ancestral proportions. Kinship was evaluated based on IBD values, removing one individual from each pair with PI_HAT>0.25 that showed a Z0, Z1, and Z2 coherent pattern (according to the theoretical expected values for each relatedness level). Genetic principal components (PCs) were calculated with PLINK with the subset of LD pruned variants.

Genotypes were imputed with the TOPMed version r2 reference panel (GRCh38) using the TOPMed Imputation Server and variants with Rsq<0.3 or with MAF<1% were filtered out. A total of 4,348 individuals and 10,671,028 genetic variants were included in the analyses.

Genetic admixture estimation

Global genetic inferred ancestry (GIA), referred to the genetic similarity to the used reference individuals, was estimated with the ADMIXTURE42 v1.3 software following a two-step procedure. First, we randomly sampled 79 European (EUR) and 79 African (AFR) samples from The 1000 Genomes Project (1KGP)43 and merged them with the 79 Native American (AMR) samples from Mao et al.44 keeping the biallelic SNPs. LD-pruned variants were selected from this merge using the same parameters as in the QC. We then run an unsupervised analysis with K=3 to redefine and homogenize the clusters and to compose a refined reference for the analyses, by applying a threshold of ≥95% of belonging to a particular cluster. As a result of this, 20 AFR, 18 EUR, and 38 AMR individuals were removed. The same LD-pruned variants data from the remaining individuals were merged with the SCOURGE Latin American cohort to perform a supervised clustering and estimated admixture proportions. A total of 471 samples from the SCOURGE cohort with >80% estimated European GIA were removed to reduce the weight of the European ancestral component, leaving a total of 3,512 admixed American (AMR) subjects for downstream analyses.

Association analysis

Results for the SCOURGE Latin Americans GWAS were obtained testing for COVID-19 hospitalization as a surrogate of severity. To accommodate the continuum of GIA in the cohort, we opted for a joint testing of all the individuals as a single study using a mixed regression model, as this approach has demonstrated a greater power and to sufficiently control population structure45. The SCOURGE cohort consisted of 3,512 COVID-19 positive patients: cases (n=1,625) were defined as hospitalized COVID-19 patients and controls (n=1,887) as non-hospitalized COVID-19 positive patients.

Logistic mixed regression models were fitted using the SAIGEgds46 package in R, which implements the two-step mixed SAIGE47 model methodology and the SPA test. Baseline covariables included sex, age, and the first 10 PCs. To account for a potential heterogeneity in the recruitment and hospitalization criteria across the participating countries, we adjusted the models by groups of the recruitment areas classified in six categories: Brazil, Colombia, Ecuador, Mexico, Paraguay, and Spain. This dataset has not been used in any previously GWAS of COVID-19 published to date.

Meta-analysis of Latin-American populations

The results of the SCOURGE Latin American cohort were meta-analyzed with the AMR HGI-B2 data, conforming our primary analysis. Summary results from the HGI freeze 7 B2 analysis corresponding to the admixed AMR population were obtained from the public repository (April 8, 2022: https://www.covid19hg.org/results/r7/), summing up 3,077 cases and 66,686 controls from seven contributing studies. We selected the B2 phenotype definition because it offered more power and the presence of population controls not ascertained for COVID-19 does not have a drastic impact in the association results.

The meta-analysis was performed using an inverse-variance weighting method in METAL48. Average allele frequency was calculated and variants with low imputation quality (Rsq<0.3) were filtered out, leaving 10,121,172 variants for meta-analysis. Heterogeneity between studies was evaluated with the Cochran’s-Q test. The inflation of results was assessed based on a genomic control (lambda).

Definition of the genetic risk loci and putative functional impact

Definition of lead variant and novel loci

To define the lead variants in the loci that were genome-wide significant, an LD-clumping was performed on the meta-analysis data using a threshold p-value<5×10-8, clump distance=1500 kb, independence set at a threshold r2=0.1 and used the SCOURGE cohort genotype data as LD reference panel. Independent loci were deemed as a novel finding if they met the following criteria: 1) p-value<5×10-8 in the meta-analysis and p-value>5×10-8 in the HGI B2 ALL meta-analysis or in the HGI B2 AMR and AFR and EUR analyses when considered by separate; 2) Cochran’s Q-test for heterogeneity of effects is <0.05/Nloci, where Nloci is the number of independent variants with p<5×10-8; and 3) the nearest gene has not been previously described in the latest HGIv7 update.

Annotation and initial mapping

Functional annotation was done with FUMA49 for those variants with a p-value<5×10-8 or in moderate-to-strong LD (r2>0.6) with the lead variants, where the LD was calculated from the 1KGP AMR panel. Genetic risk loci were defined by collapsing LD-blocks within 250 kb. Then, genes, scaled CADD v1.4 scores, and RegulomeDB v1.1 scores were annotated for the resulting variants with ANNOVAR in FUMA49. Gene-based analysis was also performed using MAGMA50 as implemented in FUMA, under the SNP-wide mean model using the 1KGP AMR reference panel. Significance was set at a threshold p<2.66×10-6 (which assumes that variants can be mapped to a total of 18,817 genes).

FUMA allowed us to perform an initial gene mapping by two approaches: (1) positional mapping, which assigns variants to genes by physical distance using 10-kb windows; and (2) eQTL mapping based on GTEx v.8 data from whole blood, lungs, lymphocytes, and oesophagus mucosa tissues, establishing a False Discovery Rate (FDR) of 0.05 to declare significance for variant-gene pairs.

Subsequently, to assign the variants to the most likely gene driving the association, we refined the candidate genes by fine mapping the discovered regions and implementing functional mapping.

To conduct a Bayesian fine mapping, credible sets for the genetic loci considered novel findings were calculated on the results from each of the three meta-analyses to identify a subset of variants most likely containing the causal variant at 95% confidence level, assuming that there is a single causal variant and that it has been tested. We used corrcoverage (https://cran.rstudio.com/web/packages/corrcoverage/index.html) for R to calculate the posterior probabilities of the variant being causal for all variants with an r2>0.1 with the leading SNP and within 1 Mb except for the novel variant in chromosome 19, for which we used a window of 0.5 Mb. Variants were added to the credible set until the sum of the posterior probabilities was ≥0.95. VEP (https://www.ensembl.org/info/docs/tools/vep/index.html) and the V2G aggregate scoring from Open Targets Genetics (https://genetics.opentargets.org) were used to annotate the biological function of the variants contained in the fine-mapped credible sets.

Colocalization analysis

We also conducted colocalization analyses to identify the putative causal genes that could act through the regulation of gene expression. FUMA’s eQTL mapping enabled the identification of genes whose expression was associated with the variants in whole blood, lungs, lymphocytes, and oesophagus mucosa tissues. We combined this information with the VEP and V2G aggregate scoring to prioritize genes. For the fine-mapping regions, we included the variants within the calculated credible sets. In the cases where the fine mapping was unsuccessful, we considered variants within a 0.2 Mb window of the lead variant.

For each prioritized gene, we then run COLOC51 to assess the evidence of colocalization between association signals and the eQTLs in each tissue, when at least one variant overlapped between them. COLOC estimates the posterior probability of two traits sharing the same causal variant in a locus. Prior probabilities of a variant being associated to COVID-19 phenotype (p1) and gene expression (p2) were set at 1×10-4, while pp2 was set at 1×10-6 as they are robust thresholds52. The posterior probability of colocalization (PP4) > 0.75 and a ratio PP4/PP3>3 were used as the criteria to support evidence of colocalization. Additionally, a threshold of PP4.SNP >0.5 was chosen for causal variant prioritization. In cases were colocalization of a single variant failed, we computed the 95% credible sets. The eQTL data was retrieved from GTEx v8 and only significant variant-gene pairs were considered in the analyses.

Colocalization in whole-blood was also performed using the recent published gene expression datasets derived from a cohort of African Americans, Puerto Ricans, and Mexican Americans (GALA II-SAGE)53. We used the results from the pooled cohort for the three discovered loci, and from the AFRHp5 (African genetic ancestry>50%) and IAMHp5 (Native American genetic ancestry>50%) cohorts for the risk loci in chromosomes 2 and 11. Results are shown in the Supplementary Table 10.

Sensitivity plots are shown in supplementary Figures 4 and 5.

Transcription-wide association studies

Transcriptome-wide association studies (TWAS) were conducted using the pretrained prediction models with MASHR-computed effect sizes on GTEx v8 datasets54,55. Results from the Latin-American meta-analysis were harmonized and integrated with the prediction models through S-PrediXcan56 for lungs, whole blood, lymphocytes and oesophagus mucosa tissues. Statistical significance was set at p-value<0.05 divided by the number of genes that were tested for each tissue. Subsequently, we leveraged results for all 49 tissues and run a multi-tissue TWAS to improve power for association, as demonstrated recently57. TWAS was also conducted with the MASHR models for whole-blood in the pooled admixed AMR from the GALA and SAGE studies53.

Cross-population meta-analyses

We conducted two additional meta-analyses to investigate the ability of combining populations to replicate our discovered risk loci. This methodology enabled the comparison of effects and the significance of associations in the novel risk loci between the results from analyses that included or excluded other population groups.

The first meta-analysis comprised the five populations analysed within HGI (B2-ALL). Additionally, to evaluate the three GIA components within the SCOURGE Latin-American cohort58, we conducted a meta-analysis of the admixed AMR, EUR, and AFR cohorts (B2). All summary statistics were retrieved from the HGI repository. We applied the same meta-analysis methodology and filters as in the admixed AMR meta-analysis. Novel variants from these meta-analyses were fine-mapped and colocalized with gene expression.

Cross-population Polygenic Risk Score

A polygenic risk score (PGS) for critical COVID-19 was derived combining the variants associated with hospitalization or disease severity that have been discovered to date. We curated a list of lead variants that were: 1) associated to either severe disease or hospitalization in the latest HGIv7 release1 (using the hospitalization weights); or 2) associated to severe disease in the latest GenOMICC meta-analysis2 that were not reported in the latest HGI release. A total of 49 markers were used in the PGS model (see supplementary Table 13) since two variants were absent from our study.

Scores were calculated and normalized for the SCOURGE Latin-American cohort with PLINK 1.9. This cross-ancestry PGS was used as a predictor for hospitalization (COVID-19 positive that were hospitalized vs. COVID-19 positive that did not necessitate hospital admission) by fitting a logistic regression model. Prediction accuracy for the PGS was assessed by performing 500 bootstrap resamples of the increase in the pseudo-R-squared. We also divided the sample in deciles and percentiles to assess risk stratification. The models were fit for the dependent variable adjusting for sex, age, the first 10 PCs, and the sampling region (in the Admixed AMR cohort) with and without the PGS, and the partial pseudo-R2 was computed and averaged among the resamples.

A clinical severity scale was used in a multinomial regression model to further evaluate the power of this cross-ancestry PGS for risk stratification. This severity strata were defined as follows: 0) asymptomatic; 1) mild, that is, with symptoms, but without pulmonary infiltrates or need of oxygen therapy; 2) moderate, that is, with pulmonary infiltrates affecting <50% of the lungs or need of supplemental oxygen therapy; 3) severe disease, that is with hospital admission and PaO2<65 mmHg or SaO2<90%, PaO2/FiO2<300, SaO2/FiO2<440, dyspnea, respiratory frequency≥22 bpm, and infiltrates affecting >50% of the lungs; and 4) critical disease, that is with an admission to the ICU or need of mechanical ventilation (invasive or non-invasive). We also included the novel risk variants as predictors alongside the PRS to determine if they provided increased prediction ability.

Data availability

Summary statistics from the SCOURGE Latin-American GWAS will be available at https://github.com/CIBERER/Scourge-COVID19.

Funding

Instituto de Salud Carlos III (COV20_00622 to A.C., COV20/00792 to M.B., COV20_00181 to C.A., COV20_1144 to M.A.J.S. and A.F.R., PI20/00876 to C.F.); European Union (ERDF) ‘A way of making Europe’. Fundación Amancio Ortega, Banco de Santander (to A.C.), Estrella de Levante S.A. and Colabora Mujer Association (to E.G.-N.) and Obra Social La Caixa (to R.B.); Agencia Estatal de Investigación (RTC-2017-6471-1 to C.F.), Cabildo Insular de Tenerife (CGIEU0000219140 ‘Apuestas científicas del ITER para colaborar en la lucha contra la COVID-19’ to C.F.) and Fundación Canaria Instituto de Investigación Sanitaria de Canarias (PIFIISC20/57 to C.F.).

SD-DA was supported by a Xunta de Galicia predoctoral fellowship.

Author contributions

Study design: RC, AC, CF. Data collection: SCOURGE cohort group. Data analysis: SD-DA, RC, ADL, CF, JML-S. Interpretation: SD-DA, RC, ADL. Drafting of the manuscript: SD-DA, RC, ADL, CF, AR-M, AC. Critical revision of the manuscript: SD-DA, RC, ADL, AC, CF, JAR, AR-M, PL. Approval of the final version of the publication: all co-authors.

Supporting information

eLife Assessment

The authors conduct a valuable GWAS meta-analysis for COVID-19 hospitalization in admixed American populations and prioritized risk variants and genes. The evidence supporting the claims of the authors is incomplete. The work will be of interest to scientists studying the genetic basis of COVID pathogenesis.

Reviewer #1 (Public Review):

Summary:

This paper conducted a GWAS meta-analysis for COVID-19 hospitalization among admixed American populations. The authors identified four genome-wide significant associations, including two novel loci (BAZ2B and DDIAS), and an additional risk locus near CREBBP using cross-ancestry meta-analysis. They utilized multiple strategies to prioritize risk variants and target genes. Finally, they constructed and assessed a polygenic risk score model with 49 variants associated with critical COVID-19 conditions.

Strengths:

Given that most of the previous studies were done in European ancestries, this study provides unique findings about the genetics of COVID-19 in admixed American populations. The GWAS data would be a valuable resource for the community. The authors conducted comprehensive analyses using multiple different strategies, including Bayesian fine mapping, colocalization, TWAS, etc., to prioritize risk variants and target genes. The polygenic risk score (PGS) result demonstrated the ability of the cross-population PGS model for COVID-19 risk stratification.

Weaknesses:

1. One of the major limitations of this study is that the GWAS sample size is relatively small, which limits its power.

2. The fine mapping section is unclear and there is a lack of information. The authors assumed one causal signal per locus, and only provided credible sets, but did not provide posterior inclusion probabilities (PIP) for the variants to be causal.

3. Colocalization and TWAS used eQTL data from GTEx data, which are mainly from European ancestries. It is unclear how much impact the ancestry mismatch would have on the result. The readers should be cautious when interpreting the results and designing follow-up studies.

Reviewer #2 (Public Review):

This is a genome-wide association study of COVID-19 in individuals of admixed American ancestry (AMR) recruited from Brazil, Colombia, Ecuador, Mexico, Paraguay, and Spain. After quality control and admixture analysis, a total of 3,512 individuals were interrogated for 10,671,028 genetic variants (genotyped + imputed). The genetic association results for these cohorts were meta-analyzed with the results from The Host Genetics Initiative (HGI), involving 3,077 cases and 66,686 controls. The authors found two novel genetic loci associated with COVID-19 at 2q24.2 (rs13003835) and 11q14.1 (rs77599934), and other two independent signals at 3p21.31 (rs35731912) and 6p21.1 (rs2477820) already reported as associated with COVID-19 in previous GWASs. Additional meta-analysis with other HGI studies also suggested risk variants near CREBBP, ZBTB7A, and CASC20 genes.

Strengths:

These findings rely on state-of-the-art methods in the field of Statistical Genomics and help to address the issue of a low number of GWASs in non-European populations, ultimately contributing to reducing health inequalities across the globe.

Weaknesses:

There is no replication cohort, as acknowledged by the authors (page 29, line 587), and no experimental validation to assess the biological effect of putative causal variants/genes. Thus, the study provides good evidence of association, rather than causation, between the genetic variants and COVID-19. Lastly, I consider it crucial to report the results for the SCOURGE Latin American GWAS, in addition to its meta-analysis with HGI results, since HGI data has a different phenotype scheme (Hospitalized COVID vs Population) compared to SCOURGE (Hospitalized COVID vs Non-hospitalized COVID).

Reviewer #3 (Public Review):

Summary:

In the context of the SCOURGE consortium's research, the authors conduct a GWAS meta-analysis on 4,702 hospitalized individuals of admixed American descent suffering from COVID-19. This study identified four significant genetic associations, including two loci initially discovered in Latin American cohorts. Furthermore, a trans-ethnic meta-analysis highlighted an additional novel risk locus in the CREBBP gene, underscoring the critical role of genetic diversity in understanding the pathogenesis of COVID-19.

Strengths:

1. The study identified two novel severe COVID-19 loci (BAZ2B and DDIAS) by the largest GWAS meta-analysis for COVID-19 hospitalization in admixed Americans.

2. With a trans-ethnic meta-analysis, an additional risk locus near CREBBP was identified.

Weaknesses:

1. The GWAS power is limited due to the relatively small number of cases.

2. There is no replication study for the novel severe COVID-19 loci, which may lead to false positive findings.

3. Significant differences exist in the ages between cases and controls, which could potentially introduce biased confounders. I'm curious about how the authors treated age as a covariate. For instance, did they use ten-year intervals? This needs clarification for reproducibility.

4."Those in the top PGS decile exhibited a 5.90-fold (95% CI=3.29-10.60, p=2.79x10-9) greater risk compared to individuals in the lowest decile". I would recommend comparing with the 40-60% PGS decile rather than the lowest decile, as the lowest PGS decile does not represent 'normal controls'.

5. In the field of PGS, it's common to require an independent dataset for training and testing the PGS model. Here, there seems to be an overfitting issue due to using the same subjects for both training and testing the variants.

6. The variants selected for the PGS appear arbitrary and may not leverage the GWAS findings without an independent training dataset.

7. The TWAS models were predominantly trained on European samples, and there is no replication study for the findings as well.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

This paper conducted a GWAS meta-analysis for COVID-19 hospitalization among admixed American populations. The authors identified four genome-wide significant associations, including two novel loci (BAZ2B and DDIAS), and an additional risk locus near CREBBP using cross-ancestry meta-analysis. They utilized multiple strategies to prioritize risk variants and target genes. Finally, they constructed and assessed a polygenic risk score model with 49 variants associated with critical COVID-19 conditions.

Strengths:

Given that most of the previous studies were done in European ancestries, this study provides unique findings about the genetics of COVID-19 in admixed American populations. The GWAS data would be a valuable resource for the community. The authors conducted comprehensive analyses using multiple different strategies, including Bayesian fine mapping, colocalization, TWAS, etc., to prioritize risk variants and target genes. The polygenic risk score (PGS) result demonstrated the ability of the cross-population

PGS model for COVID-19 risk stratification.

Thank you very much for the positive comments and the willingness to revise this manuscript.

Weaknesses:

(1) One of the major limitations of this study is that the GWAS sample size is relatively small, which limits its power.

(2) The fine mapping section is unclear and there is a lack of information. The authors assumed one causal signal per locus, and only provided credible sets, but did not provide posterior inclusion probabilities (PIP) for the variants to be causal.

(3) Colocalization and TWAS used eQTL data from GTEx data, which are mainly from European ancestries. It is unclear how much impact the ancestry mismatch would have on the result. The readers should be cautious when interpreting the results and designing follow-up studies.

We agree with that the sample size is relatively small. Despite that, it was sufficient to reveal novel risk loci supporting the robustness of the main findings. We have indicated this limitation at the end of the discussion section.

Thank you for rising this point. As suggested, we have also used SuSIE, which allows to assume more than one causal signal per locus. However, in this case the results were not different from those obtained with the original Bayesian colocalization performed with corrcoverage. Regarding the PIP, at the fine mapping stage we are inclined to put more weight on the functional annotations of the variants in the credible set than on the statistical contributions to the signal. This is the reason why we prefer not to put weight on the PIP of the variants but prioritize variants that were enriched functional annotations.

This is a good point regarding the lack of diversity in GTEx data. We have also used data from AMR populations (GALA II-SAGE models), although it was only available for blood tissue. Regarding the ancestry mismatch between datasets, several studies have attempted to explore the impact. Gay et al. (PMID: 32912333) studied local ancestry effects on eQTLs from the GTEx consortium and concluded that adjustment of eQTLs by local ancestry only yields modest improvement over using global ancestry (as done in GTEx). Moreover, the colocalization results between adjusting by Local Ancestry and Global Ancestry were not significantly different. Besides, Mogil et al. (PMID: 30096133) observed that genes with higher heritability share genetic architecture between populations. Nevertheless, both studies have evidenced decreased power and poorer predictive performances regarding gene expression because of reduced diversity in eQTL analyses. As consequence of the ancestry mismatch, we now warn the readers that this may compromise signal detection (Discussion, lines 531-533).

Reviewer #2 (Public Review):

This is a genome-wide association study of COVID-19 in individuals of admixed American ancestry (AMR) recruited from Brazil, Colombia, Ecuador, Mexico, Paraguay, and Spain. After quality control and admixture analysis, a total of 3,512 individuals were interrogated for 10,671,028 genetic variants (genotyped + imputed). The genetic association results for these cohorts were meta-analyzed with the results from The Host Genetics Initiative (HGI), involving 3,077 cases and 66,686 controls. The authors found two novel genetic loci associated with COVID-19 at 2q24.2 (rs13003835) and 11q14.1 (rs77599934), and other two independent signals at 3p21.31 (rs35731912) and 6p21.1 (rs2477820) already reported as associated with COVID-19 in previous GWASs. Additional meta-analysis with other HGI studies also suggested risk variants near CREBBP, ZBTB7A, and CASC20 genes.

Strengths:

These findings rely on state-of-the-art methods in the field of Statistical Genomics and help to address the issue of a low number of GWASs in non-European populations, ultimately contributing to reducing health inequalities across the globe.

Thank you very much for the positive comments and the willingness to revise this manuscript.

Weaknesses:

There is no replication cohort, as acknowledged by the authors (page 29, line 587), and no experimental validation to assess the biological effect of putative causal variants/genes. Thus, the study provides good evidence of association, rather than causation, between the genetic variants and COVID-19. Lastly, I consider it crucial to report the results for the SCOURGE Latin American GWAS, in addition to its meta-analysis with HGI results, since HGI data has a different phenotype scheme (Hospitalized COVID vs Population) compared to SCOURGE (Hospitalized COVID vs Non-hospitalized COVID).

We essentially agree with the reviewer in that one of the main limitations of the study is the lack of a replication stage because of the use of all available datasets on a one-stage analysis. To contribute to the interpretation of the findings in the absence of a replication stage, we now assessed the replicability of the novel loci using the Meta-Analysis Model-based Assessment of replicability (MAMBA) approach (PMID: 33785739) and included the posterior probabilities of replication in Table 2. We also explored further the potential replicability of signals in other populations. We agree that the results should be interpreted in terms of associations given the lack of functional validation of main findings, so we have slightly modified the discussion.

As suggested, the SCOURGE Latin American GWAS summary is now accessible by direct request to the Consortium GitHub repository (https://github.com/CIBERER/Scourge-COVID19) (lines 797-799). We have also included the results from the SCOURGE GWAS analysis for the replication of the 40 lead variants in the Supplementary Table 12. Results from the SCOURGE GWAS for the lead variants in the AMR meta-analysis with HGI were already included in the Supplementary Table 2. As note, we have not been able to conduct the meta-analysis with the same hospitalization scheme as in the HGI study since the population-specific results for those analyses were not publicly released. However, sensitivity analyses included within the supplementary material from the COVID-19 Host Genetics Initiative (2021) stated that there were no significant differences in effects (Odds Ratios) between analyses using population controls or just non-hospitalized COVID-19 patients.

Reviewer #3 (Public Review):

Summary:

In the context of the SCOURGE consortium's research, the authors conduct a GWAS meta-analysis on 4,702 hospitalized individuals of admixed American descent suffering from COVID-19. This study identified four significant genetic associations, including two loci initially discovered in Latin American cohorts. Furthermore, a trans-ethnic meta-analysis highlighted an additional novel risk locus in the CREBBP gene, underscoring the critical role of genetic diversity in understanding the pathogenesis of COVID-19.

Strengths:

(1) The study identified two novel severe COVID-19 loci (BAZ2B and DDIAS) by the largest GWAS meta-analysis for COVID-19 hospitalization in admixed Americans.

(2) With a trans-ethnic meta-analysis, an additional risk locus near CREBBP was identified.

Thank you very much for the positive comments and the willingness to revise this manuscript.

Weaknesses:

(1) The GWAS power is limited due to the relatively small number of cases.

(2) There is no replication study for the novel severe COVID-19 loci, which may lead to false positive findings.

We agree with that the sample size is relatively small. Despite that, it was sufficient to reveal novel risk loci supporting the robustness of the main findings. We have indicated this limitation at the end of the discussion section.

Regarding the lack of a replication study, we now assessed the replicability of the novel loci using the Meta-Analysis Model-based Assessment of replicability (MAMBA) approach (PMID: 33785739). We have included the posterior probabilities of replication in Table 2.

(3) Significant differences exist in the ages between cases and controls, which could potentially introduce biased confounders. I'm curious about how the authors treated age as a covariate. For instance, did they use ten-year intervals? This needs clarification for reproducibility.

Thank you for rising this point. Age was included as a continuous variable. This has been now indicated in line 667 (within Material and Methods).

(4)"Those in the top PGS decile exhibited a 5.90-fold (95% CI=3.29-10.60, p=2.79x10-9) greater risk compared to individuals in the lowest decile". I would recommend comparing with the 40-60% PGS decile rather than the lowest decile, as the lowest PGS decile does not represent 'normal controls'.

Thank you. In the revised version, the PGS categories was compared following the recommendation (lines 461-463).

(5) In the field of PGS, it's common to require an independent dataset for training and testing the PGS model. Here, there seems to be an overfitting issue due to using the same subjects for both training and testing the variants.

We are sorry for the misunderstanding. In fact, we have followed the standard to avoid overfitting of the PGS model and have used different training and testing datasets. The training data (GWAS) was the HGI-B2 ALL meta-analysis, in which our AMR GWAS was not included. The PRS model was then tested in the SCOURGE AMR cohort. However, it is true that we did test the combination of the PRS adding the new discovered variants in the SCOURGE cohort. To avoid potential overfitting by adding the new loci, we have excluded from the manuscript the results on which we included the newly discovered variants.

(6) The variants selected for the PGS appear arbitrary and may not leverage the GWAS findings without an independent training dataset.

Again, we are sorry for the misunderstanding. The PGS model was built with 43 variants associated with hospitalization or severity within the HGI v7 results and 7 which were discovered by the GenOMICC consortium in their latest study and were not in the latest HGI release. The variants are included within the Supplementary Table 14, but we have now annotated the discovery GWAS.

(7) The TWAS models were predominantly trained on European samples, and there is no replication study for the findings as well.

This is a good point regarding the lack of diversity in GTEx data. We have also used data from AMR populations (GALA II-SAGE models), although it was only available for blood tissue. Regarding the ancestry mismatch between datasets, several studies have attempted to explore the impact. Gay et al. (PMID: 32912333) studied local ancestry effects on eQTLs from the GTEx consortium and concluded that adjustment of eQTLs by local ancestry only yields modest improvement over using global ancestry (as done in GTEx). Moreover, the colocalization results between adjusting by Local Ancestry and Global Ancestry were not significantly different. Besides, Mogil et al. (PMID: 30096133) observed that genes with higher heritability share genetic architecture between populations. Nevertheless, both studies have evidenced decreased power and poorer predictive performances regarding gene expression because of reduced diversity in eQTL analyses. As consequence of the ancestry mismatch, we now warn the readers that this may compromise signal detection (Discussion, lines 531-533).

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

(1) The authors mentioned the fine mapping method did not converge for the locus in chr 11. I would consider trying a different fine-mapping method (such as SuSiE or FINEMAP). It would be helpful to provide posterior inclusion probabilities (PIP) for the variants in fine mapping results and plot the PIP values in the regional association plots.

As suggested, we have also used SuSIE, which allows to assume more than one causal signal per locus. However, in this case the results were not different from those obtained with the original Bayesian colocalization performed with corrcoverage. SuSIE’s fine-mapping for chromosome 11 prioritized a single variant, which is likely due to the rare frequency. Thus, we have maintained the fine-mapping as it was originally indicated in the previous version of the manuscript but have now included the credible set in Supplementary Table 6.

Regarding the PIP, at the fine mapping stage we are inclined to put more weight on the functional annotations of the variants in the credible set than on the statistical contributions to the signal. This is the reason why we prefer not to put weight on the PIP of the variants but prioritize variants that were enriched functional annotations.

(2) Please provide more detailed information about the VEP and V2G analysis and how to interpret those results. My understanding of V2G is that it includes different sources of information (such as molecular QTLs and chromatin interactions from different tissues/cell types, etc.). It is unclear what sources of information and weight settings were used in the V2G model.

Thank you for rising this point. As suggested, we have clarified the basis for VEP and V2G and the interpretation (lines 732-743).

(3) The authors identified multiple genes with different strategies, e.g. FUMA, V2G, COLOC, TWAS, etc. How many genes were found/supported by evidence provided by multiple methods? It could be helpful to have a table summarizing the risk genes found by different strategies, and the evidence supporting the genes. e.g. which genes are found by which methods, and the biological functions of the genes, etc.

Thank you for rising this point. As suggested, we now added a new figure (Figure 5) to summarize the findings with the multiple methods used.

(4) It would be helpful to make the code/scripts available for reproducibility.

As suggested, the SCOURGE Latin American GWAS summary and the analysis scripts (https://github.com/CIBERER/Scourge-COVID19/tree/main/scripts/novel-risk-hosp-AMR-2024) are now accessible in the Consortium GitHub repository (https://github.com/CIBERER/Scourge-COVID19) (lines 806-807).

(5) The fonts in some of the figures (e.g. Figure 2) are hard to read.

Thank you. We have now included the figures as SVG files.

Reviewer #2 (Recommendations For The Authors):

- The abstract lacks a conclusion sentence.

Thank you. As suggested, we have included two additional sentences with broad conclusions from the study. We preferred to avoid relying on conclusions related to known or new biological links of the prioritized genes given the lack of functional validation of main findings.

- Regarding the association analysis (page 27, line 677), I wonder if some of the 10 principal components (PCs) are capturing information about the recruitment areas (countries). It may be relevant to test for multicollinearity among these variables.

Since we acknowledge that some of the categories might be correlated with a certain PC but not all of them do, we have calculated GVIF values for the main variables to assess the categorical variable as a single entity. The scaled GVIF^1(1/2*Df)) value for the categorical variable is 1.52. Thus, if we square this value, we obtain 2.31, which can be then used for applying usual rule-of-thumb for VIF values.

- Still on the topic of association analysis, did the authors adjust the logistic model for comorbidities variables from Table 1? Given these comorbidities also have a genetic component and their distribution differs between non-hospitalized vs hospitalized, I am concerned that comorbidities might be confounding the association between genetic variants and COVID.

We did not adjust by comorbidities since HGI studies were not adjusted either and we aimed to be as aligned as possible with HGI. However, as suggested, we have now tested the association between each of the comorbidities in Table 1 and each of the variants in Table 2, using the comorbidities as dependent variables and adjusting for the main covariables (age, sex, PCs and country of recruitment). None of the variants were significantly associated to the comorbidities (line 333).

- If I understood correctly, the 49 genetic variants used to develop the polygenic risk score model (PRS) were based on the HGI total sample size (data release 7), which is predominantly of European ancestry. I am concerned about the prediction accuracy in the AMR population (PRS transferability issue).

We have explored literature in search of other PRS to compare the associated OR in our cohort with ORs calculated in European populations. Horowitz et al. (2022) reported an OR of 1.38 for the top 10% with respect to hospitalization risk in European individuals using a GRS with 12 variants.

We acknowledge that this might be an issue and is now explained in discussion of the revised version (lines 561-568). However, as this is the first time a PRS for COVID-19 is applied to a relatively large AMR cohort, we believe that this analysis will be of value for further analyses regarding PRS transferability, providing a source for comparison in further studies.

- On page 23, line 579, the authors acknowledge their "GWAS is underpowered". This sentence requires a sample/power calculation, otherwise, I suggest using "is likely underpowered".

Thanks for the input. We have modified the sentence as suggested.

Reviewer #3 (Recommendations For The Authors):

I wonder if the authors have an approximate date when the GWAS summary statistic will be available. I reviewed some manuscripts in the past, and the authors claimed they would deposit the data soon, but in fact it would not happen until 2 years later.

The summary statistics are already available from the SCOURGE Consortium repository https://github.com/CIBERER/Scourge-COVID19 (lines 806-807).

Version 2

Abstract

Abstract

The genetic basis of severe COVID-19 has been thoroughly studied, and many genetic risk factors shared between populations have been identified. However, reduced sample sizes from non-European groups have limited the discovery of population-specific common risk loci. In this second study nested in the SCOURGE consortium, we conducted a GWAS for COVID-19 hospitalization in admixed Americans, comprising a total of 4,702 hospitalized cases recruited by SCOURGE and seven other participating studies in the COVID-19 Host Genetic Initiative. We identified four genome-wide significant associations, two of which constitute novel loci and were first discovered in Latin American populations (BAZ2B and DDIAS). A trans-ethnic meta-analysis revealed another novel cross-population risk locus in CREBBP. Finally, we assessed the performance of a cross-ancestry polygenic risk score in the SCOURGE admixed American cohort. This study constitutes the largest GWAS for COVID-19 hospitalization in admixed Latin Americans conducted to date. This allowed to reveal novel risk loci and emphasize the need of considering the diversity of populations in genomic research.

Introduction

To date, more than 50 loci associated with COVID-19 susceptibility, hospitalization, and severity have been identified using genome-wide association studies (GWAS)1,2. The COVID-19 Host Genetics Initiative (HGI) has made significant efforts3 to augment the power to identify disease loci by recruiting individuals from diverse populations and conducting a trans-ancestry meta-analysis. Despite this, the lack of genetic diversity and a focus on cases of European ancestries still predominate in the studies4,5. In addition, while trans-ancestry meta-analyses are a powerful approach for discovering shared genetic risk variants with similar effects across populations6, they may fail to identify risk variants that have larger effects on particular underrepresented populations. Genetic disease risk has been shaped by the particular evolutionary history of populations and environmental exposures7. Their action is particularly important for infectious diseases due to the selective constraints that are imposed by host–pathogen interactions8,9. Literature examples of this in COVID-19 severity include a DOCK2 gene variant in East Asians10 and frequent loss-of-function variants in IFNAR1 and IFNAR2 genes in Polynesian and Inuit populations, respectively11,12.

Including diverse populations in case–control GWAS studies with unrelated participants usually requires a prior classification of individuals in genetically homogeneous groups, which are typically analyzed separately to control the population stratification effects13. Populations with recent admixture impose an additional challenge to GWASs due to their complex genetic diversity and linkage disequilibrium (LD) patterns, requiring the development of alternative approaches and a careful inspection of results to reduce false positives due to population structure7. In fact, there are benefits in study power from modeling the admixed ancestries either locally, at the regional scale in the chromosomes, or globally, across the genome, depending on factors such as the heterogeneity of the risk variant in frequencies or the effects among the ancestry strata14. Despite the development of novel methods specifically tailored for the analysis of admixed populations15, the lack of a standardized analysis framework and the difficulties in confidently clustering admixed individuals into particular genetic groups often leads to their exclusion from GWAS.

The Spanish Coalition to Unlock Research on Host Genetics on COVID-19 (SCOURGE) recruited COVID-19 patients between March and December 2020 from hospitals across Spain and from March 2020 to July 2021 in Latin America (https://www.scourge-covid.org). A first GWAS of COVID-19 severity among Spanish patients of European descent revealed novel disease loci and explored age- and sex-varying effects of genetic factors16. Here, we present the findings of a GWAS meta-analysis in admixed Latin American (AMR) populations, comprising individuals from the SCOURGE Latin-American cohort and the HGI studies, which allowed us to identify two novel severe COVID-19 loci, BAZ2B and DDIAS. Further analyses modeling the admixture from three genetic ancestral components and performing a trans-ethnic meta-analysis led to the identification of an additional risk locus near CREBBP. We finally assessed a cross-ancestry polygenic risk score model with variants associated with critical COVID-19.

Results

Meta-analysis of COVID-19 hospitalization in admixed Americans

Study cohorts

Within the SCOURGE consortium, we included 1,608 hospitalized cases and 1,887 controls (not hospitalized COVID-19 patients) from Latin American countries and from recruitments of individuals of Latin American descent conducted in Spain (Supplementary Table 1). Quality control details and estimation of global genetic inferred ancestry (GIA) (Supplementary Figure 1) are described in Methods, whereas clinical and demographic characteristics of patients included in the analysis are shown in Table 1. Summary statistics from the SCOURGE cohort were obtained under a logistic mixed model with the SAIGE model (Methods). Another seven studies participating in the COVID-19 HGI consortium were included in the meta-analysis of COVID-19 hospitalization in admixed Americans (Figure 1).

Flow chart of this study.

GWAS meta-analysis

We performed a fixed-effects GWAS meta-analysis using the inverse of the variance as weights for the overlapping markers. The combined GWAS sample size consisted of 4,702 admixed AMR hospitalized cases and 68,573 controls.

This GWAS meta-analysis revealed genome-wide significant associations at four risk loci (Table 2, Figure 2), two of which (BAZ2B and DDIAS) were novel discoveries. Four lead variants were identified, linked to other 310 variants (Supplementary Tables 2-3). A gene-based association test revealed a significant association in BAZ2B and in previously known COVID-19 risk loci: LZTFL1, XCR1, FYCO1, CCR9, and IFNAR2 (Supplementary Table 4).

A) Manhattan plot for the admixed AMR GWAS meta-analysis. Probability thresholds at p=5×10−8 and p=5×10−5 are indicated by the horizontal lines. Genome-wide significant associations with COVID-19 hospitalizations were found on chromosome 2 (within BAZ2B), chromosome 3 (within LZTFL1), chromosome 6 (within FOXP4), and chromosome 11 (within DDIAS). A Quantile–Quantile plot is shown in supplementary Figure 2. B) Regional association plots for rs1003835 at chromosome 2 and rs77599934 at chromosome 11; C) Allele frequency distribution across the 1000 Genomes Project populations for the lead variants rs1003835 and rs77599934. Retrieved from The Geography of Genetic Variants Web or GGV.

Located within the BAZ2B gene, the sentinel variant rs13003835 is an intronic variant associated with an increased risk of COVID-19 hospitalization (Odds Ratio [OR]=1.20, 95% Confidence Interval [CI]=1.12-1.27, p=3.66×10−8). This association was not previously reported in any GWAS of COVID-19 published to date. Interestingly, rs13003835 did not reach significance (p=0.972) in the COVID-19 HGI trans-ancestry meta-analysis including the five population groups1.

The other novel risk locus is led by the sentinel variant rs77599934, a rare intronic variant located in chromosome 11 within DDIAS and associated with the risk of COVID-19 hospitalization (OR=2.27, 95% CI=1.70-3.04, p=2.26×10−8).

We also observed a suggestive association with rs2601183 in chromosome 15, which is located between ZNF774 and IQGAP1 (allele-G OR=1.20, 95% CI=1.12-1.29, p=6.11×10−8, see Supplementary Table 2), which has not yet been reported in any other GWAS of COVID-19 to date.

The GWAS meta-analysis also pinpointed two significant variants at known loci, LZTFL1 and FOXP4. The SNP rs35731912 was previously associated with COVID-19 severity in EUR populations17, and it was mapped to LZTFL1. While rs2477820 is a novel risk variant within the FOXP4 gene, it has a moderate LD (r2=0.295) with rs2496644, which has been linked to COVID-19 hospitalization18. This is consistent with the effects of LD in tag-SNPs when conducting GWAS in diverse populations.

None of the lead variants was associated with the comorbidities included in Table 1.

Functional mapping of novel risk variants

Variants belonging to the lead loci were prioritized by positional and expression quantitative trait loci (eQTL) mapping with FUMA, resulting in 31 mapped genes (Supplementary Table 5). Within the region surrounding the lead variant in chromosome 2, FUMA prioritized four genes in addition to BAZ2B (PLA2R1, LY75, WDSUB1, and CD302). rs13003835 (allele C) is an eQTL of LY75 in the esophagus mucosa (NES=0.27) and of BAZ2B-AS in whole blood (NES=0.27), while rs2884110 (R2=0.85) is an eQTL of LY75 in lung (NES=0.22). As for the chromosome 11, rs77599934 (allele G) is in moderate-to-strong LD (r2=0.776) with rs60606421 (G deletion, allele -) which is an eQTL associated with a reduced expression of DDIAS in the lungs (NES=-0.49, allele -). Associations with expression are shown in the supplementary Figure 4. The sentinel variant for the region in chromosome 16 is in perfect LD (r2=1) with rs601183, an eQTL of ZNF774 in the lung.

Bayesian fine mapping

We performed different approaches to narrow down the prioritized loci to a set of most likely genes driving the associations. First, we computed credible sets at the 95% confidence level for causal variants and annotated them with VEP (and V2G aggregate scoring. The 95% confidence credible set from the region of chromosome 2 around rs13003835 included 76 variants, which can be found in the Supplementary Table 6 (VEP and V2G annotations are included in the Supplementary Tables 7 and 8). TheV2G score prioritized BAZ2B as the most likely gene driving the association. However, the approach was unable to converge allocating variants in a 95% confidence credible set for the region in chromosome 11.

Transcriptome-wide association study (TWAS)

Five novel genes, namely, SLC25A37, SMARCC1, CAMP, TYW3, and S100A12 (supplementary Table 9), were found to be significantly associated in the cross-tissue TWAS. To our knowledge, these genes have not been reported previously in any COVID-19 TWAS or GWAS analyses published to date. In the single tissue analyses, ATP5O and CXCR6 were significantly associated in the lungs, CCR9 was significantly associated in whole blood, and IFNAR2 and SLC25A37 were associated in lymphocytes.

Likewise, we carried out TWAS analyses using the models trained in the admixed populations. However, no significant gene pairs were detected in this case. The top 10 genes with the lowest p-values for each of the datasets (Puerto-Ricans, Mexicans, African-Americans and pooled cohorts) are shown in the Supplementary Table 10. Although not significant, KCNC3 was repeated in the four analyses, whereas MAPKAPK3, NAPSA and THAP5 were repeated in 3 out of 4. Both NAPSA and KCNC3 are located in the chromosome 19 and were reported in the latest HGI meta-analysis19.

All mapped genes from analyses conducted in AMR populations are shown in Figure 5.

Genetic architecture of COVID-19 hospitalization in AMR populations

Allele frequencies of rs13003835 and rs77599934 across ancestries

Neither rs13003835 (BAZ2B) nor rs77599934 (DDIAS) were significantly associated in the COVID-19 HGI B2 cross-population or population-specific meta-analyses. Thus, we investigated their allele frequencies (AF) across populations and compared their effect sizes.

According to gnomAD v3.1.2, the T allele at rs13003835 (BAZ2B) has an AF of 43% in admixed AMR groups, while AF is lower in the EUR populations (16%) and in the global sample (29%). Local ancestry inference (LAI) reported by gnomAD shows that within the Native-American component, the risk allele T is the major allele, whereas it is the minor allele within the African and European LAI components. These large differences in AF might be the reason underlying the association found in AMR populations. However, when comparing effect sizes between populations, we found that they were in opposite directions between SAS-AMR and EUR-AFR-EAS and that there was large heterogeneity among them (Figure 3). We queried SNPs within 50kb windows of the lead variant in each of the other populations that had p-value<0.01. The variant with the lowest p-value in the EUR population was rs559179177 (p=1.72×10−4), which is in perfect LD (r2=1) in the 1KGP EUR population with our sentinel variant (rs13003835), and in moderate LD r2=0.4 in AMR populations. Since this variant was absent from the AMR analysis, probably due to its low frequency, it could not be meta-analyzed. Power calculations revealed that the EUR analysis was underpowered for this variant to achieve genome-wide significance (77.6%, assuming an effect size of 0.46, EAF= 0.0027, and number of cases/controls as shown in the HGI website for B2-EUR). In the cross-population meta-analysis (B2-ALL), rs559179177 obtained a p-value of 5.9×10−4.

Forest plot showing effect sizes and the corresponding confidence intervals for the sentinel variants identified in the AMR meta-analysis across populations.

All beta values with their corresponding CIs were retrieved from the B2 population-specific meta-analysis from the HGI v7 release, except for AMR, for which the beta value and IC from the HGIAMR-SCOURGE meta-analysis are represented.

rs77599934 (DDIAS) had an AF of 1.1% for the G allele in the nonhospitalized controls (Table 2), in line with the recorded gnomAD AF of 1% in admixed AMR groups. This variant has the potential to be a population-specific variant, given the allele frequencies in other population groups, such as EUR (0% in Finnish, 0.025% in non-Finnish), EAS (0%) and SAS (0.042%), and its greater effect size over AFR populations (Figure 3). Examining the LAI, the G allele occurs at a 10.8% frequency in the African component, while it is almost absent in the Native-American and European. Due to its low MAF, rs77599934 was not analyzed in the COVID-19 HGI B2 cross-population meta-analysis and was only present in the HGI B2 AFR population-specific meta-analysis, precluding the comparison (Figure 3). For this reason, we retrieved the variant with the lowest p-value within a 50 kb region around rs77599934 in the COVID-19 HGI cross-population analysis to investigate whether it was in moderate-to-strong LD with our sentinel variant. The variant with the smallest p-value was rs75684040 (OR=1.07, 95% CI=1.03-1.12, p=1.84×10−3). However, LD calculations using the 1KGP phase 3 dataset indicated that rs77599934 and rs75684040 were poorly correlated (r2=0.11). As for AFR populations, the variant with the lowest p-value was rs138860115 (p=8.3×10−3), but it was not correlated with the lead SNP of this locus.

Cross-population meta-analyses

We carried out two cross-ancestry inverse variance-weighted fixed-effects meta-analyses with the admixed AMR GWAS meta-analysis results to evaluate whether the discovered risk loci replicated when considering other population groups. In doing so, we also identified novel cross-population COVID-19 hospitalization risk loci.

First, we combined the SCOURGE Latin American GWAS results with the HGI B2 ALL analysis (supplementary Table 11). We refer to this analysis as the SC-HGIALL meta-analysis. Out of the 40 genome-wide significant loci associated with COVID-19 hospitalization in the last HGI release1, this study replicated 39, and the association was stronger than in the original study in 29 of those (supplementary Table 12). However, the variant rs13003835 located in BAZ2B did not replicate (OR=1.00, 95% CI=0.98-1.03, p=0.644).

In this cross-ancestry meta-analysis, we replicated two associations that were not found in HGIv7, albeit they were sentinel variants in the latest GenOMICC meta-analysis2. We found an association at the CASC20 locus led by the variant rs2876034 (OR=0.95, 95% CI=0.93-0.97, p=2.83×10−8). This variant is in strong LD with the sentinel variant of that study (rs2326788, r2=0.92), which was associated with critical COVID-192. In addition, this meta-analysis identified the variant rs66833742 near ZBTB7A associated with COVID-19 hospitalization (OR=0.94, 95% CI=0.92-0.96, p=2.50×10−8). Notably, rs66833742 or its perfect proxy rs67602344 (r2=1) are also associated with upregulation of ZBTB7A in whole blood and in esophageal mucosa. This variant was previously associated with COVID-19 hospitalization2.

In a second analysis, we also explored the associations across the defined admixed AMR, EUR, and AFR ancestral sources by combining through meta-analysis the SCOURGE Latin American GWAS results with the HGI studies in EUR, AFR, and admixed AMR and excluding those from EAS and SAS (supplementary Table 13). We refer to this as the SC-HGI3POP meta-analysis. The association at rs13003835 (BAZ2B, OR=1.01, 95% CI=0.98-1.03, p=0.605) was not replicated, and rs77599934 near DDIAS could not be assessed, although the association at the ZBTB7A locus was confirmed (rs66833742, OR=0.94, 95% CI=0.92-0.96, p=1.89×10−8). The variant rs76564172 located near CREBBP also reached statistical significance (OR=1.31, 95% CI=1.25-1.38, p=9.64×10−9). The sentinel variant of the region linked to CREBBP (in the trans-ancestry meta-analysis) was also subjected a Bayesian fine mapping (supplementary Table 6). Eight variants were included in the credible set for the region in chromosome 16 (meta-analysis SC-HGI3POP).

Polygenic risk score models

Using the 49 variants associated with disease severity that are shared across populations according to the HGIv7, we constructed a polygenic risk score (PGS) model to assess its generalizability in the admixed AMR (supplementary Table 14). First, we calculated the PGS for the SCOURGE Latin Americans and explored the association with COVID-19 hospitalization under a logistic regression model. The PGS model was associated with a 1.48-fold increase in COVID-19 hospitalization risk per every PGS standard deviation. It also contributed to explaining a slightly larger variance (ΔR2=1.07%) than the baseline model. Subsequently, we divided the individuals into PGS deciles and percentiles to assess their risk stratification. The median percentile among controls was 40, while in cases, it was 63. Those in the top PGS decile exhibited a 2.89-fold (95% CI=2.37-3.54, p=1.29×10−7) greater risk compared to individuals in the deciles between 4 and 6 (corresponding to a score of the median distribution).

We also examined the distribution of PGS scores across a 5-level severity scale to further determine if there was any correspondence between clinical severity and genetic risk. Median PGS scores were lower in the asymptomatic and mild groups, whereas higher median scores were observed in the moderate, severe, and critical patients (Figure 4). We fitted a multinomial model using the asymptomatic class as a reference and calculated the OR for each category (supplementary Table 15), observing that the disease genetic risk was similar among asymptomatic, mild, and moderate patients. Given that the PGS was built with variants associated with critical disease and/or hospitalization and that the categories severe and critical correspond to hospitalized patients, these results underscore the ability of cross-ancestry PGS for risk stratification even in an admixed population.

(A) Polygenic risk stratified by PGS deciles comparing each risk group against the lowest risk group (OR-95% CI); (B) Distribution of the PGS scores in each of the severity scale classes.

0-Asymptomatic, 1-Mild disease, 2-Moderate disease, 3-Severe disease, 4-Critical disease.

Summary of the results from gene prioritization strategies used for genetic associations in AMR populations.

GWAS catalog association for BAZ2B-AS was with FEV/FCV ratio. Literature based evidence is further explored in discussion.

Discussion

We have conducted the largest GWAS meta-analysis of COVID-19 hospitalization in admixed AMR to date. While the genetic risk basis discovered for COVID-19 is largely shared among populations, trans-ancestry meta-analyses on this disease have primarily included EUR samples. This dominance of GWAS in Europeans and the subsequent bias in sample sizes can mask population-specific genetic risks (i.e., variants that are monomorphic in some populations) or be less powered to detect risk variants having higher allele frequencies in population groups other than Europeans. In this sense, after combining data from admixed AMR patients, we found two risk loci that were first discovered in a GWAS of Latin American populations. Interestingly, the sentinel variant rs77599934 in the DDIAS gene is a rare coding variant (∼1% for allele G) with a large effect on COVID-19 hospitalization that is nearly monomorphic in most of the other populations. This has likely led to its exclusion from the cross-population meta-analyses conducted to date, remaining undetectable.

Fine mapping of the region harboring DDIAS did not reveal further information about which gene could be the more prone to be causal or about the functional consequences of the risk variant, but our sentinel variant was in strong LD with an eQTL that associated with reduced gene expression of DDIAS in the lung. DDIAS, known as damage-induced apoptosis suppressor gene, is itself a plausible candidate gene. It has been linked to DNA damage repair mechanisms: research has shown that depletion of DDIAS leads to an increase in ATM phosphorylation and the formation of p53-binding protein (53BP1) foci, a known biomarker of DNA double-strand breaks, suggesting a potential role in double-strand break repair20. Interestingly, a study found that infection by SARS-CoV-2 also triggered the phosphorylation of ATM kinase and inhibited repair mechanisms, causing the accumulation of DNA damage21. This gene has also been proposed as a potential biomarker for lung cancer after finding that it interacts with STAT3 in lung cancer cells, regulating IL-622,23 and thus mediating inflammatory processes, while another study determined that its blockade inhibited lung cancer cell growth24.Another prioritized gene from this region was PRCP, an angiotensinase which shares substrate specificity with ACE2 receptor. It has been positively linked to hypertension and some studies have raised hypotheses on its role in COVID-19 progression, particularly in relation to the development of pro-thrombotic events 25,26.

The risk region found in chromosome 2 harbors more than one gene. The lead variant rs13003835 is located within BAZ2B, and it increases the expression of the antisense BAZ2B gene in whole blood. BAZ2B encodes one of the regulatory subunits of the Imitation switch (ISWI) chromatin remodelers27 constituting the BRF-1/BRF-5 complexes with SMARCA1 and SMARCA5, respectively. Interestingly, it was discovered that lnc-BAZ2B promotes macrophage activation through regulation of BAZ2B expression. Its over-expression resulted in pulmonary inflammation and elevated levels of MUC5AC in mice with asthma28. This variant was also an eQTL for LY75 (encoding lymphocyte antigen 75) in the esophageal mucosa tissue. Lymphocyte antigen 75 is involved in immune processes through antigen presentation in dendritic cells and endocytosis29 and has been associated with inflammatory diseases, representing a compelling candidate for the region. Increased expression of LY75 has been detected within hours after infection by SARS-CoV-230,31. It is worth noting that differences in AF for this variant suggest that analyses in AMR populations might be more powered to detect the association, supporting the necessity of population-specific studies.

A third novel risk region was observed on chromosome 15 between the IQGAP1 and ZNF774 genes, although it did not reach genome-wide significance.

Secondary analyses revealed five TWAS-associated genes, some of which have already been linked to severe COVID-19. In a comprehensive multitissue gene expression profiling study32, decreased expression of CAMP and S100A8/S100A9 genes in patients with severe COVID-19 was observed, while another study detected the upregulation of SCL25A37 among patients with severe COVID-1933. SMARCC1 is a subunit of the SWI/SNF chromatin remodeling complex that has been identified as proviral for SARS-CoV-2 and other coronavirus strains through a genome-wide screen34. This complex is crucial for ACE2 expression and viral entry into the cell35. However, it should be noted that using eQTL mostly from European populations such as those in GTEx could result in reduced power to detect associations.

To explore the genetic architecture of the trait among admixed AMR populations, we performed two cross-ancestry meta-analyses including the SCOURGE Latin American cohort GWAS findings. We found that the two novel risk variants were not associated with COVID-19 hospitalization outside the population-specific meta-analysis, highlighting the importance of complementing trans-ancestry meta-analyses with group-specific analyses. Notably, this analysis did not replicate the association at the DSTYK locus, which was associated with severe COVID-19 in Brazilian individuals with higher European admixture36. This lack of replication aligns with the initial hypothesis of that study suggesting that the risk haplotype was derived from European populations, as we reduced the weight of this ancestral contribution in our study by excluding those individuals.

Moreover, these cross-ancestry meta-analyses pointed to three loci that were not genome-wide significant in the HGIv7 ALL meta-analysis: a novel locus at CREBBP and two loci at ZBTB7A and CASC20 that were reported in another meta-analysis. CREBBP and ZBTB7A achieved a stronger significance when considering only the EUR, AFR, and admixed AMR GIA groups. According to a recent study, elevated levels of the ZBTB7A gene promote a quasihomeostatic state between coronaviruses and host cells, preventing cell death by regulating oxidative stress pathways37. This gene is involved in several signaling pathways, such as B and T-cell differentiation38. On a separate note, CREBBP encodes the CREB binding protein (CBP), which is involved in transcriptional activation and is known to positively regulate the type I interferon response through virus-induced phosphorylation of IRF-339. In addition, the CREBP/CBP interaction has been implicated in SARS-CoV-2 infection40 via the cAMP/PKA pathway. In fact, cells with suppressed CREBBP gene expression exhibit reduced replication of the so-called Delta and Omicron SARS-CoV-2 variants40.

We developed a cross-population PGS model, which effectively stratified individuals based on their genetic risk and demonstrated consistency with the clinical severity classification of the patients. Only a few polygenic scores were derived from COVID-19 GWAS data. Horowitz et al. (2022)41 developed a score using 6 and 12 associated variants (PGS ID: PGP000302) and reported an associated OR (top 10% vs rest) of 1.38 for risk of hospitalization in European populations, whereas the OR for Latin-American populations was 1.56. Since their sample size and the number of variants included in the PGS were lower, direct comparisons are not straightforward. Nevertheless, our analysis provides the first results for a PRS applied to a relatively large AMR cohort, being of value for future analyses regarding PRS transferability.

This study is subject to limitations, mostly concerning sample recruitment and composition. The SCOURGE Latino-American sample size is small, and the GWAS is likely underpowered. Another limitation is the difference in case–control recruitment across sampling regions that, yet controlled for, may reduce the ability to observe significant associations driven by different compositions of the populations. In this sense, the identified risk loci might not replicate in a cohort lacking any of the parental population sources from the three-way admixture. Likewise, we could not explicitly control for socioenvironmental factors that could have affected COVID-19 spread and hospitalization rates, although genetic principal components are known to capture nongenetic factors. Finally, we must acknowledge the lack of a replication cohort. We used all the available GWAS data for COVID-19 hospitalization in admixed AMR in this meta-analysis due to the low number of studies conducted. Therefore, we had no studies to replicate or validate the results. These concerns may be addressed in the future by including more AMR GWAS in the meta-analysis, both by involving diverse populations in study designs and by supporting research from countries in Latin America.

This study provides novel insights into the genetic basis of COVID-19 severity, emphasizing the importance of considering host genetic factors by using non-European populations, especially of admixed sources. Such complementary efforts can pin down new variants and increase our knowledge on the host genetic factors of severe COVID-19.

Materials and methods

GWAS in Latin Americans from SCOURGE

The SCOURGE Latin American cohort

A total of 3,729 COVID-19-positive cases were recruited across five countries from Latin America (Mexico, Brazil, Colombia, Paraguay, and Ecuador) by 13 participating centers (supplementary Table 1) from March 2020 to July 2021. In addition, we included 1,082 COVID-19-positive individuals recruited between March and December 2020 in Spain who either had evidence of origin from a Latin American country or showed inferred genetic admixture between AMR, EUR, and AFR (with < 0.05% SAS/EAS). These individuals were excluded from a previous SCOURGE study that focused on participants with European genetic ancestries16. We used hospitalization as a proxy for disease severity and defined COVID-19-positive patients who underwent hospitalization as a consequence of the infection as cases and those who did not need hospitalization due to COVID-19 as controls.

Samples and data were collected with informed consent after the approval of the Ethics and Scientific Committees from the participating centers and by the Galician Ethics Committee Ref 2020/197. Recruitment of patients from IMSS (in Mexico, City) was approved by the National Committee of Clinical Research from Instituto Mexicano del Seguro Social, Mexico (protocol R-2020-785-082).

Samples and data were processed following normalized procedures. The REDCap electronic data capture tool42,43, hosted at Centro de Investigación Biomédica en Red (CIBER) from the Instituto de Salud Carlos III (ISCIII), was used to collect and manage demographic, epidemiological, and clinical variables. Subjects were diagnosed with COVID-19 based on quantitative PCR tests (79.3%) or according to clinical (2.2%) or laboratory procedures (antibody tests: 16.3%; other microbiological tests: 2.2%).

SNP array genotyping

Genomic DNA was obtained from peripheral blood and isolated using the Chemagic DNA Blood 100 kit (PerkinElmer Chemagen Technologies GmbH), following the manufacturer’s recommendations.

Samples were genotyped with the Axiom Spain Biobank Array (Thermo Fisher Scientific) following the manufacturer’s instructions in the Santiago de Compostela Node of the National Genotyping Center (CeGen-ISCIII; http://www.usc.es/cegen). This array contains probes for genotyping a total of 757,836 SNPs. Clustering and genotype calling were performed using Axiom Analysis Suite v4.0.3.3 software.

Quality control steps and variant imputation

A quality control (QC) procedure using PLINK 1.944 was applied to both samples and the genotyped SNPs. We excluded variants with a minor allele frequency (MAF) <1%, a call rate <98%, and markers strongly deviating from Hardy-Weinberg equilibrium expectations (p<1×10−6) with mid-p adjustment. We also explored the excess of heterozygosity to discard potential cross-sample contamination. Samples missing >2% of the variants were filtered out. Subsequently, we kept the autosomal SNPs, removed high-LD regions and conducted LD pruning (windows of 1,000 SNPs, with a step size of 80 and a r2 threshold of 0.1) to assess kinship and estimate the global ancestral proportions. Kinship was evaluated based on IBD values, removing one individual from each pair with PI_HAT>0.25 that showed a Z0, Z1, and Z2 coherent pattern (according to the theoretical expected values for each relatedness level). Genetic principal components (PCs) were calculated with PLINK with the subset of LD pruned variants.

Genotypes were imputed with the TOPMed version r2 reference panel (GRCh38) using the TOPMed Imputation Server, and variants with Rsq<0.3 or with MAF<1% were filtered out. A total of 4,348 individuals and 10,671,028 genetic variants were included in the analyses.

Genetic admixture estimation

Global genetic inferred ancestry (GIA), referred to the genetic similarity to the used reference individuals, was estimated with ADMIXTURE45 v1.3 software following a two-step procedure. First, we randomly sampled 79 European (EUR) and 79 African (AFR) samples from The 1000 Genomes Project (1KGP)46 and merged them with the 79 Native American (AMR) samples from Mao et al.47 keeping the biallelic SNPs. LD-pruned variants were selected from this merge using the same parameters as in the QC. We then run an unsupervised analysis with K=3 to redefine and homogenize the clusters and to compose a refined reference for the analyses by applying a threshold of ≥95% of belonging to a particular cluster. As a result, 20 AFR, 18 EUR, and 38 AMR individuals were removed. The same LD-pruned variants data from the remaining individuals were merged with the SCOURGE Latin American cohort to perform supervised clustering and estimate admixture proportions. A total of 471 samples from the SCOURGE cohort with >80% estimated European GIA were removed to reduce the weight of the European ancestral component, leaving a total of 3,512 admixed Latin American (AMR) subjects for downstream analyses.

Association analysis

The results for the SCOURGE Latin American GWAS were obtained by testing for COVID-19 hospitalization as a surrogate of severity. To accommodate the continuum of GIA in the cohort, we opted for a joint testing of all the individuals as a single study using a mixed regression model, as this approach has demonstrated a greater power and sufficient control of population structure48. The SCOURGE cohort consisted of 3,512 COVID-19-positive patients: cases (n=1,625) were defined as hospitalized COVID-19 patients, and controls (n=1,887) were defined as nonhospitalized COVID-19-positive patients.

Logistic mixed regression models were fitted using the SAIGEgds49 package in R, which implements the two-step mixed SAIGE50 model methodology and the SPA test. Baseline covariables included sex, age (continuous), and the first 10 PCs. To account for potential heterogeneity in the recruitment and hospitalization criteria across the participating countries, we adjusted the models by groups of the recruitment areas classified into six categories: Brazil, Colombia, Ecuador, Mexico, Paraguay, and Spain. This dataset has not been used in any previously published GWAS of COVID-19.

Meta-analysis of Latin American populations

The results of the SCOURGE Latin American cohort were meta-analyzed with the AMR HGI-B2 data, conforming our primary analysis. Summary results from the HGI freeze 7 B2 analysis corresponding to the admixed AMR population were obtained from the public repository (April 8, 2022: https://www.covid19hg.org/results/r7/), summing up 3,077 cases and 66,686 controls from seven contributing studies. We selected the B2 phenotype definition because it offered more power, and the presence of population controls not ascertained for COVID-19 does not have a drastic impact on the association results.

The meta-analysis was performed using an inverse-variance weighting method in METAL51. The average allele frequency was calculated, and variants with low imputation quality (Rsq<0.3) were filtered out, leaving 10,121,172 variants for meta-analysis.

Heterogeneity between studies was evaluated with Cochran’s Q test. The inflation of results was assessed based on a genomic control (lambda).

Replicability of associations

The model-based method MAMBA52 was used to calculate the posterior probabilities of replication for each of the lead variants. Variants with p<1×10−05 were clumped and combined with random pruned variants from the 1KGP AMR reference panel. Then, MAMBA was applied to the set of significant and non-significant variants.

Each of the lead variants was also tested for association with the main comorbidities in the SCOURGE cohort with logistic regression models (adjusted by the same base covariables as the GWAS).

Definition of the genetic risk loci and putative functional impact

Definition of lead variant and novel loci

To define the lead variants in the loci that were genome-wide significant, LD-clumping was performed on the meta-analysis data using a threshold p-value<5×10−8, clump distance=1500 kb, independence set at a threshold r2=0.1 and the SCOURGE cohort genotype data as the LD reference panel. Independent loci were deemed as a novel finding if they met the following criteria: 1) p-value<5×10−8 in the meta-analysis and p-value>5×10−8 in the HGI B2 ALL meta-analysis or in the HGI B2 AMR and AFR and EUR analyses when considered separately; 2) Cochran’s Q-test for heterogeneity of effects is <0.05/Nloci, where Nloci is the number of independent variants with p<5×10−8; and 3) the nearest gene has not been previously described in the latest HGIv7 update.

Annotation and initial mapping

Functional annotation was performed with FUMA53 for those variants with a p-value<5×10−8 or in moderate-to-strong LD (r2>0.6) with the lead variants, where the LD was calculated from the 1KGP AMR panel. Genetic risk loci were defined by collapsing LD blocks within 250 kb. Then, genes, scaled CADD v1.4 scores, and RegulomeDB v1.1 scores were annotated for the resulting variants with ANNOVAR in FUMA53. Gene-based analysis was also performed using MAGMA54 as implemented in FUMA under the SNP-wide mean model using the 1KGP AMR reference panel. Significance was set at a threshold p<2.66×10−6 (which assumes that variants can be mapped to a total of 18,817 genes).

FUMA allowed us to perform initial gene mapping by two approaches: (1) positional mapping, which assigns variants to genes by physical distance using 10-kb windows; and (2) eQTL mapping based on GTEx v.8 data from whole blood, lungs, lymphocytes, and esophageal mucosa tissues, establishing a false discovery rate (FDR) of 0.05 to declare significance for variant-gene pairs.

Subsequently, to assign the variants to the most likely gene driving the association, we refined the candidate genes by fine mapping the discovered regions

Bayesian fine-mapping

To conduct a Bayesian fine mapping, credible sets for the genetic loci considered novel findings were calculated on the results from each of the three meta-analyses to identify a subset of variants most likely containing the causal variant at the 95% confidence level, assuming that there is a single causal variant and that it has been tested. We used corrcoverage (https://cran.rstudio.com/web/packages/corrcoverage/index.html) for R to calculate the posterior probabilities of the variant being causal for all variants with a r2>0.1 with the leading SNP and within 1 Mb except for the novel variant in chromosome 19, for which we used a window of 0.5 Mb. Variants were added to the credible set until the sum of the posterior probabilities was ≥0.95.

VEP and V2G annotation

We used the Variant-to-Gene (V2G) score to prioritize the genes that were most likely affected by the functional evidence based on expression quantitative trait loci (eQTL), chromatin interactions, in silico functional predictions, and distance between the prioritized variants and transcription start site (TSS), based on data from the Open Targets Genetics portal55. Details of the data integration and the weighting of each of the datasets are described with detail here: https://genetics-docs.opentargets.org/our-approach/data-pipeline. V2G is a score for ranking the functional genomics evidence that supports the connection between variants and genes (the higher the score the more likely the variant to be functionally implicated on the assigned gene). We used VEP release 111 (URL: https://www.ensembl.org/info/docs/tools/vep/index.html; accessed April 10, 2024) 56 to annotate the following: gene symbol, function (exonic, intronic, intergenic, non-coding RNA, etc.), impact, feature type, feature, and biotype.

We queried the GWAS Catalog (date of accession: 01/07/2024) for evidence of association of each of the prioritized genes with traits related to lung diseases or phenotypes. Lastly, those which were linked to COVID-19, infection, or lung diseases in the revised literature were classified as “literature evidence”.

Transcription-wide association studies

Transcriptome-wide association studies (TWAS) were conducted using the pretrained prediction models with MASHR-computed effect sizes on GTEx v8 datasets57,58. The results from the Latin American meta-analysis were harmonized and integrated with the prediction models through S-PrediXcan59 for lung, whole blood, lymphocyte and esophageal mucosal tissues. Statistical significance was set at p-value<0.05 divided by the number of genes that were tested for each tissue. Subsequently, we leveraged results for all 49 tissues and ran a multitissue TWAS to improve the power for association, as demonstrated recently60. TWAS was also performed using recently published gene expression datasets derived from a cohort of African Americans, Puerto Ricans, and Mexican Americans (GALA II-SAGE)61.

Cross-population meta-analyses

We conducted two additional meta-analyses to investigate the ability of combining populations to replicate our discovered risk loci. This methodology enabled the comparison of effects and the significance of associations in the novel risk loci between the results from analyses that included or excluded other population groups.

The first meta-analysis comprised the five populations analyzed within HGI (B2-ALL). Additionally, to evaluate the three GIA components within the SCOURGE Latin American cohort62, we conducted a meta-analysis of the admixed AMR, EUR, and AFR cohorts (B2). All summary statistics were retrieved from the HGI repository. We applied the same meta-analysis methodology and filters as in the admixed AMR meta-analysis.

Cross-population Polygenic Risk Score

A polygenic risk score (PGS) for critical COVID-19 was derived by combining the variants associated with hospitalization or disease severity that have been discovered to date. We curated a list of lead variants that were 1) associated with either severe disease or hospitalization in the latest HGIv7 release1 (using the hospitalization weights) or 2) associated with severe disease in the latest GenOMICC meta-analysis2 that were not reported in the latest HGI release. A total of 48 markers were used in the PGS model (see supplementary Table 13) since two variants were absent from our study.

Scores were calculated and normalized for the SCOURGE Latin American cohort with PLINK 1.9. This cross-ancestry PGS was used as a predictor for hospitalization (COVID-19-positive patients who were hospitalized vs. COVID-19-positive patients who did not necessitate hospital admission) by fitting a logistic regression model. Prediction accuracy for the PGS was assessed by performing 500 bootstrap resamples of the increase in the pseudo-R-squared. We also divided the sample into deciles and percentiles to assess risk stratification. The models were fit for the dependent variable adjusting for sex, age, the first 10 PCs, and the sampling region (in the Admixed AMR cohort) with and without the PGS, and the partial pseudo-R2 was computed and averaged among the resamples.

A clinical severity scale was used in a multinomial regression model to further evaluate the power of this cross-ancestry PGS for risk stratification. These severity strata were defined as follows: 0) asymptomatic; 1) mild, that is, with symptoms, but without pulmonary infiltrates or need of oxygen therapy; 2) moderate, that is, with pulmonary infiltrates affecting <50% of the lungs or need of supplemental oxygen therapy; 3) severe disease, that is, with hospital admission and PaO2<65 mmHg or SaO2<90%, PaO2/FiO2<300, SaO2/FiO2<440, dyspnea, respiratory frequency≥22 bpm, and infiltrates affecting >50% of the lungs; and 4) critical disease, that is, with an admission to the ICU or need of mechanical ventilation (invasive or noninvasive).